An algorithm for finding the global extremum of a partially defined function\thanks{This work was supported by the Russian Science Foundation, project No. 21-11-00204.}

### 2023-08-28 2023/MMST/samplepaper.tex
add last paragraph, fix typos and fix to 15 pages

First, the objective function may be multiextremal, nondifferentiable, and, moreo-ver, defined in the form of a "black box"\ (i. e. in the form of some subroutine, which entry the argument is supplied to, and the output is respective function value).

@@

Third, the objective function satisfies the Lipschitz condition
\begin{equation}\label{eq3} 
| \phi (y')-\phi (y'') | \leq L \| y'-y'' \|, y',y' \in D,
\end{equation}
where $L$ is the Lipschitz constant. The dimensionality reduction scheme using Peano curves is known to juxtapose the multidimensional problem with a Lipschitz objective function (\ref{eq1}) to problem (\ref{eq2}) with the one-dimensional objective function satisfying H{\"o}lder condition
\begin{equation}\label{eq4} 
| \phi (x')-\phi (x'') | \leq K \| x'-x'' \| ^{1/N}, x',x' \in [0,1],
\end{equation}
where $N$ is the dimensionality of the initial multidimensional problem and coefficient $K$ is related to the Lipschitz constant $L$ by the relation $K \leq 4L\sqrt N$ \cite{Strongin2000}.

@@

It is worth noting that some machine learning algorithms can take into account and even restore the missing values in the data. For example, LightGBM has a regime of ignoring the missing values while XGBoost restores the data by means of the de-crease in the loss function when learning. Nevertheless, the majority of models can-not process the missing values and require the valid data without NaN or «missing» values. It makes this problem to be serious enough.

@@

        \paragraph{Case 1} For the intervals with \textbf{two computable points} or with \textbf{one computable point and one boundary point}, apply the classic rules of global search algorithm
\begin{equation}\label{eq21} 
R(i)=
  \begin{cases}
     \Delta _i+\frac {{(z_i-z_{i-1})}^2}{{(r \mu)}^2 \Delta _i} - 2 \frac {z_i+z_{i-1}-2z_1^*}{r \mu}, & {\quad  v(x_i)=v(x_{i-1})=1}\\
    2 \Delta _i-4 \frac {(z_i-z_1^*)}{r \mu}, & {\quad  v(x_{i-1})=-1, v(x_i)=1}\\
    2 \Delta _i-4 \frac {(z_{i-1}-z_1^*)}{r \mu}, & {\quad  v(x_{i-1})=1, v(x_i)=-1}
  \end{cases}
\end{equation}
        \paragraph{Case 2}   For the intervals with \textbf{one boundary point and one non-computable point}, perform the imputation of the values for the non-computable point based on the neighbor computable point and apply the formula
\begin{equation}\label{eq22} 
R(i)=
  \begin{cases}
    2 \Delta _i-4 \frac {(z_{i+1}-z_0^*)}{r \mu}, & {\quad  v(x_{i-1})=-1, v(x_i)=0}\\
    2 \Delta _i-4 \frac {(z_{i-2}-z_0^*)}{r \mu}, & {\quad  v(x_{i-1})=0, v(x_i)=-1}
  \end{cases}
\end{equation}
        \paragraph{Case 3} For the intervals with \textbf{one non-computable and one computable point}, perform the imputation of the value for the non-computable point based on the neighbor com-putable point and apply the formula
\begin{equation}\label{eq23} 
R(i)=
  \begin{cases}
     \Delta _i+\frac {{(z_i-z_{i-2})}^2}{{(r \mu)}^2 \Delta _i} - 2 \frac {z_i+z_{i-2}-2z_0^*}{r \mu}, & {\quad  v(x_{i-1})=0, v(x_i)=1}\\
     \Delta _i+\frac {{(z_{i+1}-z_{i-1})}^2}{{(r \mu)}^2 \Delta _i} - 2 \frac {z_{i+1}+z_{i-1}-2z_0^*}{r \mu}, & {\quad  v(x_{i-1})=1, v(x_i)=0}
  \end{cases}
\end{equation}
        \paragraph{Case 4} For the intervals with \textbf{two non-computable points}, perform the imputation of the values for the non-computable points based on the rule (\ref{eq19}) and apply the formula
\begin{equation}\label{eq24} 
R(i)=\Delta _i+\frac {{(\tilde{z}_i-\tilde{z}_{i-1})}^2}{{(r \mu)}^2 \Delta _i} - 2 \frac {\tilde{z}_i+\tilde{z}_{i-1}-2z_1^*}{r \mu}, v(x_{i-1})=v(x_i)=0
\end{equation}
        \paragraph{Case 5} In the case when any imputation is impossible, use the formula
\begin{equation}\label{eq25} 
R(i)=\Delta _i \cdot {\left( 1-\frac {1}{r} \right)}^N+z_0^*.
\end{equation}
  \item If several intervals have equal characteristics, prefer the computable interval with the lowest number.
  \item If the best interval appears to be the one with two non-computable points, perform the next trial in the middle point of the interval.
\end{enumerate}

@@

\subsection{Conclusions}
In the present work, the problem of finding the global minimum of a computation-costly "black box"\ function was considered. As compared to the traditional formulation used in the global optimization, the problem considered has an important differ-ence: the objective function may be undefined in some regions of the search domain (in particular, in one point or in several ones). The appearing of such a problem statement is caused by a frequent problem of numerical instability of the models investigated in the applied problems.

@@

[....rus...] $N = 2,3,4,5$ ...

>>>>

First, the objective function may be multiextremal, nondifferentiable, and, moreover, defined in the form of a "black box"\ (i. e. in the form of some subroutine, which entry the argument is supplied to, and the output is respective function value).

@@

Third, the objective function satisfies the Lipschitz condition
\begin{equation}\label{eq3} 
| \phi (y')-\phi (y'') | \leq L \| y'-y'' \|, y',y' \in D,
\end{equation}
where $L$ is the Lipschitz constant. The dimensionality reduction scheme using Peano curves is known to juxtapose the multidimensional problem with a Lipschitz objective function (\ref{eq1}) to problem (\ref{eq2}) with the one-dimensional objective function satisfying H{\"o}lder condition
\begin{equation}\label{eq4} 
| \phi (x')-\phi (x'') | \leq K \| x'-x'' \| ^{1/N}, x',x' \in [0,1],
\end{equation}
where $N$ is the dimensionality of the initial multidimensional problem and coefficient $K$ is related to the Lipschitz constant $L$ by the rule $K \leq 4L\sqrt N$ \cite{Strongin2000}.

@@

It is worth noting that some machine learning algorithms can take into account and even restore the missing values in the data. For example, LightGBM has a regime of ignoring the missing values while XGBoost restores the data by means of the decrease in the loss function when learning. Nevertheless, the majority of models cannot process the missing values and require the valid data without NaN or «missing» values. It makes this problem to be serious enough.

@@

        \paragraph{Case 1} For the intervals with \textbf{two computable points} or with \textbf{one computable point and one boundary point}, apply the classic rules of global search algorithm
\begin{equation}\label{eq21} 
R(i)=
  \begin{cases}
     \Delta _i+\frac {{(z_i-z_{i-1})}^2}{{(r \mu)}^2 \Delta _i} - 2 \frac {z_i+z_{i-1}-2z_1^*}{r \mu}, & {\quad  v(x_i)=v(x_{i-1})=1}\\
    2 \Delta _i-4 \frac {(z_i-z_1^*)}{r \mu}, & {\quad  v(x_{i-1})=-1, v(x_i)=1}\\
    2 \Delta _i-4 \frac {(z_{i-1}-z_1^*)}{r \mu}, & {\quad  v(x_{i-1})=1, v(x_i)=-1}
  \end{cases}
\end{equation}
        \paragraph{Case 2}   For the intervals with \textbf{one boundary point and one non-computable point}, perform the imputation of the values for the non-computable point based on the neighbor computable point and apply the formula
\begin{equation}\label{eq22} 
R(i)=
  \begin{cases}
    2 \Delta _i-4 \frac {(z_{i+1}-z_0^*)}{r \mu}, & {\quad  v(x_{i-1})=-1, v(x_i)=0}\\
    2 \Delta _i-4 \frac {(z_{i-2}-z_0^*)}{r \mu}, & {\quad  v(x_{i-1})=0, v(x_i)=-1}
  \end{cases}
\end{equation}
        \paragraph{Case 3} For the intervals with \textbf{one non-computable and one computable point}, perform the imputation of the value for the non-computable point based on the neighbor computable point and apply the formula
\begin{equation}\label{eq23} 
R(i)=
  \begin{cases}
     \Delta _i+\frac {{(z_i-z_{i-2})}^2}{{(r \mu)}^2 \Delta _i} - 2 \frac {z_i+z_{i-2}-2z_0^*}{r \mu}, & {\quad  v(x_{i-1})=0, v(x_i)=1}\\
     \Delta _i+\frac {{(z_{i+1}-z_{i-1})}^2}{{(r \mu)}^2 \Delta _i} - 2 \frac {z_{i+1}+z_{i-1}-2z_0^*}{r \mu}, & {\quad  v(x_{i-1})=1, v(x_i)=0}
  \end{cases}
\end{equation}
        \paragraph{Case 4} For the intervals with \textbf{two non-computable points}, perform the imputation of the values for the non-computable points based on the rule (\ref{eq19}) and apply the formula
\begin{equation}\label{eq24} 
R(i)=\Delta _i+\frac {{(\tilde{z}_i-\tilde{z}_{i-1})}^2}{{(r \mu)}^2 \Delta _i} - 2 \frac {\tilde{z}_i+\tilde{z}_{i-1}-2z_1^*}{r \mu}, v(x_{i-1})=v(x_i)=0
\end{equation}
        \paragraph{Case 5} In the case when any imputation is impossible, use the formula
\begin{equation}\label{eq25} 
R(i)=\Delta _i \cdot {\left( 1-\frac {1}{r} \right)}^N+z_0^*.
\end{equation}
  \item If several intervals have equal characteristics, prefer the computable interval with the lowest number.
  \item If the best interval appears to be the one with two non-computable points, perform the next trial in the middle point of the interval.
\end{enumerate}

@@

\subsection{Conclusions}
In the present work, the problem of finding the global minimum of a computation-costly "black box"\ function was considered. As compared to the traditional formulation used in the global optimization, the problem considered has an important difference: the objective function may be undefined in some regions of the search domain (in particular, in one point or in several ones). The appearing of such a problem statement is caused by a frequent problem of numerical instability of the models investigated in the applied problems.

@@

The numerical experiments were carried out using series of multiextremal GKLS problems of the dimensionalities $N = 2,3,4,5$. The results have demonstrated the reliability and efficiency of the approach proposed. The modified algorithm handles successfully the problems, which the non-computability regions are scattered randomly within the search domain. In the case of large boundary non-computability regions, the method solves the problems efficiently as well performing a complete investigation of the non-computability regions also.
